---
author: ""
categories: ["R"]
tags: ["tidyverse", "ggplot2"]
date: "2020-01-27"
description: "Quick Reference Guide"
featured: "ggplot2.png"
featuredalt: "Six Things I Always Google When Using ggplot2"
featuredpath: "date"
linktitle: ""
title: "Six Things I Always Google When Using ggplot2"
type: "post"
---

I often use {ggplot2} to create graphs but there are certain things I *always* have to Google. I figured I'd create a post for quick reference for myself but I'd love to hear what you always have to look up!

* [Remove the Legend](#remove-the-legend)
* [Change Legend Title and Labels](#change-legend-title-and-labels)
* [Manually Change Colors](#manually-change-colors)
* [Remove X Axis Labels](#remove-x-axis-labels)
* [Start the Y Axis at a Specific Number](#start-the-y-axis-at-a-specific-number)
* [Use Scales on the Y Axis](#use-scales-on-the-y-axis)

```{r, message = F, warning = F}
library(tidyverse)

knitr::opts_chunk$set(out.width = '100%') 
```

To showcase what's happening, I am going to use a [TidyTuesday](https://github.com/rfordatascience/tidytuesday) dataset: Spotify songs! Let's start by creating a simple graph.

```{r, message = F, warning = F}
# Load Data
spotify_songs <- 
  readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')

spotify_songs %>% 
  ggplot(aes(x = playlist_genre)) +
  geom_histogram(stat = "count")
```

## Remove the legend

>  theme(legend.position = "none")

Ahh... this one always gets me. Sometimes when your color is mostly just for aesthetics,  it doesn't make sense to also have a color legend. This removes the legend and makes the graph look cleaner.

```{r, warning = F, message = F}
spotify_songs %>% 
  ggplot(aes(x = playlist_genre, fill = playlist_genre)) +
  geom_histogram(stat = "count") +
  theme(legend.position = "none")
```

## Change Legend Title and Labels

> scale_fill_discrete(name = "New Legend Title", labels = c("lab1" = "Label 1", "lab2" = "Label 2"))

Alright, say I do want the legend. How do I make it something readable?

```{r, warning = F, message = F}
spotify_songs %>% 
  ggplot(aes(x = playlist_genre, fill = playlist_genre)) +
  geom_histogram(stat = "count") +
  scale_fill_discrete(name = "Playlist Genre", 
                      labels = c("edm" = "EDM", 
                                 "latin" = "Latin", 
                                 "pop" = "Pop", 
                                 "r&b" = "R&B", 
                                 "rap" = "Rap", 
                                 "rock" = "Rock"))
```

## Manually Change Colors

> scale_fill_manual("New Legend Title", values = c("lab1" = "#000000", "lab2" = "#FFFFFF"))

This is a bit tricker, in that you cannot use `scale_fill_manual` and `scale_fill_discrete` separately on the same plot as they override each other. However, if you want to change the labels *and* the colors together, you can use `scale_fill_manual` like below.

```{r, warning = F, message = F}
spotify_songs %>% 
  ggplot(aes(x = playlist_genre, fill = playlist_genre)) +
  geom_histogram(stat = "count") +
  scale_fill_manual(name = "Playlist Genre", 
                    labels = c("edm" = "EDM", 
                               "latin" = "Latin", 
                               "pop" = "Pop", 
                               "r&b" = "R&B", 
                               "rap" = "Rap", 
                               "rock" = "Rock"),
                    values = c("edm" = "#68B39B", 
                               "latin" = "#F6C7FF", 
                               "pop" = "#ADFFE5", 
                               "r&b" = "#CCB576", 
                               "rap" = "#B3A070", 
                               "rock" = "#d3d3d3"))
```

## Remove X Axis Labels

>  theme(axis.title.x = element_blank(),
         axis.text.x = element_blank(),
         axis.ticks.x = element_blank())
        
In this case, since we have a legend, we don't need any x axis labels. Sometimes I use this if there's redundant information or if it otherwise makes the graph look cleaner.

```{r, warning = F, message = F}
spotify_songs %>% 
  ggplot(aes(x = playlist_genre, fill = playlist_genre)) +
  geom_histogram(stat = "count") +
  scale_fill_manual(name = "Playlist Genre", 
                    labels = c("edm" = "EDM", 
                               "latin" = "Latin", 
                               "pop" = "Pop", 
                               "r&b" = "R&B", 
                               "rap" = "Rap", 
                               "rock" = "Rock"),
                    values = c("edm" = "#68B39B", 
                               "latin" = "#F6C7FF", 
                               "pop" = "#ADFFE5", 
                               "r&b" = "#CCB576", 
                               "rap" = "#B3A070", 
                               "rock" = "#d3d3d3")) +
  theme(axis.title.x = element_blank(),
         axis.text.x = element_blank(),
         axis.ticks.x = element_blank())
```

## Start the Y Axis at a Specific Number

> scale_y_continuous(name = "New Y Axis Title", limits = c(0, 1000000))

Often times, we want our graph's y axis to start at 0. In this example it already does, but this handy parameter allows us to set exactly what we want our y axis to be.

```{r, warning = F, message = F}
spotify_songs %>% 
  ggplot(aes(x = playlist_genre, fill = playlist_genre)) +
  geom_histogram(stat = "count") +
  scale_fill_manual(name = "Playlist Genre", 
                    labels = c("edm" = "EDM", 
                               "latin" = "Latin", 
                               "pop" = "Pop", 
                               "r&b" = "R&B", 
                               "rap" = "Rap", 
                               "rock" = "Rock"),
                    values = c("edm" = "#68B39B", 
                               "latin" = "#F6C7FF", 
                               "pop" = "#ADFFE5", 
                               "r&b" = "#CCB576", 
                               "rap" = "#B3A070", 
                               "rock" = "#d3d3d3")) +
  theme(axis.title.x = element_blank(),
         axis.text.x = element_blank(),
         axis.ticks.x = element_blank()) +
  scale_y_continuous(name = "Count", limits = c(0, 10000))
```

## Use scales on the Y Axis

> scale_y_continuous(label = scales::format)

Depending on our data, we may want the y axis to be formatted a certain way (using dollar signs, commas, percentage signs, etc.). The handy {scales} package allows us to do that easily. 

```{r, warning = F, message = F}
spotify_songs %>% 
  ggplot(aes(x = playlist_genre, fill = playlist_genre)) +
  geom_histogram(stat = "count") +
  scale_fill_manual(name = "Playlist Genre", 
                    labels = c("edm" = "EDM", 
                               "latin" = "Latin", 
                               "pop" = "Pop", 
                               "r&b" = "R&B", 
                               "rap" = "Rap", 
                               "rock" = "Rock"),
                    values = c("edm" = "#68B39B", 
                               "latin" = "#F6C7FF", 
                               "pop" = "#ADFFE5", 
                               "r&b" = "#CCB576", 
                               "rap" = "#B3A070", 
                               "rock" = "#d3d3d3")) +
  theme(axis.title.x = element_blank(),
         axis.text.x = element_blank(),
         axis.ticks.x = element_blank()) +
  scale_y_continuous(name = "Count", limits = c(0, 10000),
                     labels = scales::comma)
```

There we have it! Six things I always eventually end up Googling when I am making plots using {ggplot2}. Hopefully now I can just look at this page instead of searching each and every time!
